---
title: "Reproducible Research Week 2 Peer-graded Assignment: Course Project 1"
author: "Robert J.Coon"
date: "6/16/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Project Assignment

## Loading and Preprocessing the Data

###    Show any code that is needed to load and, if necessary, process/transform the data into a format suitable for your analysis

1. Load the data.

```{r}
setwd("C:\\Users\\bcoon\\Desktop\\Files\\Coursera\\")

the_data <- read.table(unz(".\\Data\\repdata_data_activity.zip",
                           "activity.csv" ), header = T, sep = ",")

summary(the_data)
```


### What is the mean number of steps taken per day?
1.  Calculate the total number of steps per day:  
  (Aggregate the data by date to determine the total number of steps taken per day)

```{r}
the_agg_data <- aggregate(the_data$steps, by=list(Category = the_data$date), FUN=sum)

the_agg_data

```

2.   Make a histogram of the total number of steps taken each day (Chose ggplot to generate this histogram):

```{r}
library("ggplot2")

steps_per_day_hist <- ggplot(the_agg_data, aes(x)) + geom_histogram(binwidth = 2500, 
                              col = "red", fill="blue") + 
                              labs(x = "Steps", y = "Number of Occurrences",
                              title = "Histogram of Steps per Day")

steps_per_day_hist

```

3.  Calculate and report the mean and median of the total number of steps taken per day:  

##  Mean steps per day: 

```{r}
the_total_mean <- mean(the_agg_data$x, na.rm = TRUE)
the_total_mean
```

##  Median steps per day: 

```{r}
the_total_median <- median(the_agg_data$x, na.rm = TRUE)
the_total_median
```

#  Average Daily Activity Pattern

1. Time series plot of the 5 minute interval (x) and averaged number of steps taken averaged across all days (y)

```{r}

the_intervals_agg <- aggregate(the_data$steps, list(the_data$interval), FUN = mean, na.rm = TRUE )

the_line_plot <- ggplot(the_intervals_agg, aes(x = Group.1, y = x)) + geom_line() +
                        labs(title = "Steps Taken per Time Interval", subtitle = 
                            "Time Series of Average Steps", y = "Number of Steps", 
                             x = "Time Interval")

the_line_plot

```


2.  Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}

max_interval <- the_intervals_agg[which(the_intervals_agg$x == max(the_intervals_agg$x)),]

max_interval

```

#  Imputing Missing Values

To do this, create a dataset with missing data ("NA" values) excluded, create a table of the aggregated means, and then fill in the missing values in the original dataset with the means from the aggregated dataset matched on the corresponding intervals.

1.  Calculate and report the total number of missing values in the dataset:
To do this, simply calculate the number of rows with missing values for the "steps" column value:

```{r}
num_na_rows <- nrow(the_data[the_data$steps == "NA", ])

num_na_rows

```

2.  Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.
3.  Create a new dataset that is equal to the original dataset but with the missing data filled in.
To fulfill steps 2. and 3., create a dataset with missing data ("NA" values) excluded, create a table of the aggregated means, and then fill in the missing values in the original dataset with the means from the aggregated dataset matched on the corresponding intervals.

```{r}
the_new_data <- na.omit(the_data)
the_new_agg_data <- aggregate(the_new_data$steps, list(the_new_data$interval), mean)
the_final_data <- the_data
the_final_data$step_count <- ifelse(is.na(the_final_data$steps), the_new_agg_data$x[match(the_final_data$interval, 
                                     the_new_agg_data$Group.1)], the_final_data$steps)
the_final_data$steps <- round(the_final_data$step_count,0)
the_final_agg_data <- aggregate(the_final_data$steps, list(the_final_data$date), sum)
str(the_final_data)
```

4.  Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. 

```{r}
imputed_data_hist <- ggplot(the_final_agg_data, aes(x)) + geom_histogram(binwidth = 2500, col = "red", fill="blue")

imputed_data_hist <- imputed_data_hist + labs(x = "Steps", y = "Number of Occurrences",
                              title = "Histogram of Steps per Day",
                              subtitle = "(Missing Data Values Imputed)")

imputed_data_hist

```

##  Mean steps per day using imputed md data: 

```{r}
the_total_md_mean <- mean(the_final_agg_data$x, na.rm = TRUE)
the_total_md_mean
```

##  Median steps per day using imputed md data: 

```{r}
the_total_md_median <- median(the_final_agg_data$x, na.rm = TRUE)
the_total_md_median
```

#  Weekend and Weekday Activity Differences:

1.  Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.


```{r}

summary(the_final_data)
the_final_data$day_type <- weekdays(as.Date(the_final_data$date))

the_final_data$day_type <- ifelse(the_final_data$day_type == "Saturday" | the_final_data$day_type == "Sunday",
                                    "Weekend", "Weekday")
```

2.  Make a panel plot containing a time series plot (i.e. \color{red}{\verb|type = "l"|}type="l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). 

```{r}
library("ggplot2")

the_daytype_agg <- aggregate(step_count ~ interval + day_type, data = the_final_data, 
                             FUN = mean, na.action = na.omit)

summary(the_daytype_agg)

the_daytype_line_plot <- ggplot(the_daytype_agg, aes(x = interval, y = step_count, col = "red")) + 
    geom_line() +
    labs(title = "Steps Taken per Time Interval", subtitle = 
         "Time Series of Average Steps by Weekday vs. Weekend", y = "Number of Steps", 
       x = "Time Interval") + facet_grid(day_type ~ .) + theme(legend.position = "none")

the_daytype_line_plot
```











